---
title: "Introduction to time series analysis"
author: "Boyko Amarov"
date: "10/12/2021"
output: html_document
---

# Introduction to time series analysis

## Some examples of time series

- temperature (weather), e.g. daily average temperature
- Rainfall series
- Prices of stocks/debt
- GDP series
- Inflation series


```{r}
# install.packages("xts")
# Extended Time Series
library(tidyverse)
library(xts)

bitcoins <- read_csv(
  'https://raw.githubusercontent.com/feb-sofia/timeseries-2022-2023/main/data/bitcoins.csv')
```


- Plot the time series

```{r}
# Create a time series object with xts
bc_ts <- xts(bitcoins$Close, order.by = as.Date(bitcoins$Date))["2017/"]
```


```{r}
plot(bc_ts)
```


Homework: provide a brief description of the main patterns visible in the time
series plot

- Compute the average exchange rate over the whole observation period

```{r}
## Using the mean function

mean(bc_ts)
```
- Compute the average exchange rate for June 2017

```{r}
## Subset the time series, then apply the mean function
mean(bc_ts["2017-6"])
```

- Compute the average exchange rate for October 2017

```{r}
## Same as for June 2017
mean(bc_ts["2017-10"])
```

- Compute the average exchange rate before (and including) and after October 2017

```{r}
mean(bc_ts["/2017-10"])
```
```{r}
mean(bc_ts["2017-11/"])
```
- Compute the standard deviation for the four periods

What does the standard deviation measure?
It measures how much the daily exchange rates vary around their mean: volatility (variation).

```{r}
## Subset the time series, then apply the sd function (which computes the standard deviation)
sd(bc_ts["2017-6"])
```

```{r}
## Same as for June 2017
sd(bc_ts["2017-10"])
```

```{r}
sd(bc_ts["/2017-10"])
```

```{r}
sd(bc_ts["2017-11/"])
```

- Compute the average exchange rate for every week during the observation period

```{r}
bc_weekly_avg <- apply.weekly(bc_ts, mean)
bc_weekly_avg
plot(bc_weekly_avg)
```


- Compute the standard deviation of the exchange rate for every week during the observation period

```{r}
bc_weekly_sd <- apply.weekly(bc_ts, sd)
bc_weekly_sd
plot(bc_weekly_sd)
```
