---
title: "Основни понятия"
author: "Boyko Amarov"
date: "10/12/2021"
output: html_document
---

```{r setup}
# install.packages("xts")
# Extended Time Series
library(tidyverse)
library(xts)

bitcoins <- read_csv(
  'https://raw.githubusercontent.com/feb-sofia/timeseries-2022-2023/main/data/bitcoins.csv')
bc_ts <- xts(bitcoins$Close, order.by = as.Date(bitcoins$Date))["2017/"]
```

# The World of Data

## Mathematical notation

$$
y_{t}, t = 1,\ldots,T: \text{ denotes a time series with T observations}\\
y_1, y_2, y_3, \ldots, y_{T}
$$
where $T$ is the index of the last observation. Look at the `bc_ts` object to determine the following:  $T = ?$. $y_1 = ?, y_2 = ?, \ldots, y_{?} = ?$.


Empirical (sample) average (mean)

$$
\hat{\mu} = \frac{1}{T}\sum_{t = 1}^{T} y_{t}
$$

Empirical (sample) variance

$$
\hat{\gamma}_{0} = \frac{1}{T}\sum_{t = 1}^{T}(y_t - \hat{\mu})^2 \text{ variance}\\
\sqrt{\hat{\gamma}_0} \text{ standard deviation}
$$

Exercise: compute the sample mean and variance of the following series:

$$
y_1 = 2, y_2 = 3, y_3 = 10
$$


```{r}
mean(bc_ts)
var(bc_ts)
```

## Lags

Looking from Nov 13th 2017: what was the value of the exchange rate on the previous trading day? Answer
by simply looking in the table: $y_{117 - 1} = y_{?} = ?$.

$$
y_{t}, t = 1,\ldots, T\\
y_{t - 1}: \text{ first lag of } y_{t}\\
y_{t - 2}: \text{ second lag of } y_{t}
$$

Exercise: write down the first two lag series of

$$
y_1 = 2, y_2 = 3, y_3 = 10
$$

Construct the lag 1 and lag two series in the bitcoins example.

```{r prep-lags}
bc_ts_l1 <- stats::lag(bc_ts, k = 1)
bc_ts_l2 <- stats::lag(bc_ts, k = 2)
bc_ts_l3 <- stats::lag(bc_ts, k = 3)

bc_ts_lags <- tibble(
  y = bc_ts,
  y_l1 = bc_ts_l1,
  y_l2 = bc_ts_l2,
  y_l3 = bc_ts_l3
)

bc_ts_lags
```

Create a scatterplot for the exchange rate time series and its first lag

```{r plot-lag-1}
lags_scatter <- bc_ts_lags %>%
  ggplot(aes(x = y_l1, y = y)) +
  geom_point() +
  labs(
    title = "Scatterplot of y[t] and y[t-1]",
    x = expression(y[t - 1]),
    y = expression(y[t])
  )
lags_scatter
```

Compute the empirical mean and variance of the first lag series in the bitcoins example:

```{r}
## Watch out for errors due to missing values
mean(bc_ts_l1)
((length(bc_ts) - 1) / length(bc_ts)) * var(bc_ts_l1)
```

### Empirical auto-covariances and auto-correlations

$$
\hat{\gamma}_{1} = \frac{1}{T}\sum_{t = 2}^{T} (y_{t} - \hat{\mu})(y_{t - 1} - \hat{\mu})\\
\rho_{1} = \frac{\hat{\gamma}_1}{\hat{\gamma}_0}
$$

The sign of this sum in the covariance definition depends on the arrangement of the points relative to the
series average.

```{r}
lags_scatter +
  ## Add horizontal line at the average exchange rate
  geom_hline(yintercept = mean(bc_ts), color = "steelblue2") +
  ## Add vertical line at the average exchange rate
  geom_vline(xintercept = mean(bc_ts), color = "steelblue2")
```

$$
(y_t - \hat{\mu}) \text{ the deviation of y_t from the average}\\
(y_{t - 1} - \hat{\mu}) \text{ the deviation of first lag from the average}\\
(y_{t} - \hat{\mu})(y_{t - 1} - \hat{\mu}) \text{ cross product of deviations}
$$

$$
\text{Cross product of deviations from the means}\\
(y_{t} - )(y_i - \hat{\mu})\\
(x_i - \hat{\mu})(y_i - \hat{\mu}) \geq 0 \text{ for days in the upper right part of the plot: high-high}\\
(x_i - \hat{\mu})(y_i - \hat{\mu}) \geq 0 \text{ for days in the lower left of the plot: low-low}\\
(x_i - \hat{\mu})(y_i - \hat{\mu}) \leq 0 \text{ for days in the upper left of the plot: low-high}\\
(x_i - \hat{\mu})(y_i - \hat{\mu}) \leq 0 \text{ for days in the upper left of the plot: high-low}
$$

$$
\sum_{i=1}^{T}(y_i - \hat{\mu})(y_i - \hat{\mu}) = \\
\sum_{i \in \text{low-low}}\underbrace{(x_i - \hat{\mu})(y_i - \hat{\mu})}_{+} +
\sum_{i \in \text{high-high}}\underbrace{(x_i - \hat{\mu})(y_i - \hat{\mu})}_{+} +\\
\sum_{i \in \text{low-high}}\underbrace{(x_i - \hat{\mu})(y_i - \hat{\mu})}_{-} +
\sum_{i \in \text{high-low}}\underbrace{(x_i - \hat{\mu})(y_i - \hat{\mu})}_{-}
$$

## Simulation: a random series

In the present simulation we will use the `rnorm` function to draw a series of $n = 100$ values from the standard normal distribution
with $\mu = 0$ and $\sigma = 1$: $N(0, 1)$.

```{r}
## Draw n = 100 values from the standard normal distribution
set.seed(123)
x <- rnorm(n = 100, mean = 0, sd = 1)

ggplot() +
  stat_function(data = tibble(x = c(-3, 3)), aes(x = x), fun = dnorm, n = 101, args = list(mean = 0, sd = 1)) +
  ylab("") +
  scale_y_continuous(breaks = NULL) +
  geom_rug(data = tibble(x = x), aes(x = x))
```

```{r}
ts_random <- ts(x)
plot(ts_random)
```


```{r}
ts_random_l1 <- stats::lag(ts_random)
ts_random_lag <- tibble(
  ts_random,
  ts_random_l1
)

ts_random_lag %>%
  ggplot(aes(x = ts_random_l1, y = ts_random)) +
  geom_point() +
  ggtitle('') +
  ylab('y[t]') +
  xlab('y[t - 1]') +
  geom_hline(yintercept = mean(ts_random), color = "steelblue2") +
  geom_vline(xintercept = mean(ts_random), color = "steelblue2")
```

```{r}
stats::lag(x)
```

# World of mathematics

We will use the same notation as for the observed values, so you should rely on the context to
distinguish whether we are referring to observed values or to random variables.

$$
y_1, y_2, \ldots, y_T
$$

Mathematical expectation (expected value, mean)

$$
E(y_t)
$$

this will describe the average of the values generated from $y_t$.

$$
Var(y_t) = E\(y_t - Ey_{t}\)^2
$$

Variance: this will describe how different the values generated from $y_t$ are.

$$
Cov(y_t, y_{t - 1}) = E(y_{t} - Ey_{t})(y_{t - 1} - Ey_{t - 1})
$$

Covariance: this will describe how the different generator are related.


$$
\hat{\mu} = \frac{2 + 3 + 10}{3} = 5
$$
$$
S^2_{y} = \frac{1}{3}\left((y_1 - \hat{\mu})^2 + (y_2 - \hat{\mu})^2 + (y_3 - \hat{\mu})^2 \right)=\\
S^2_{y} = \frac{1}{3}\left((2 - 5)^2 + (3 - 5)^2 + (10 - 5)^2 \right)=\\
S^2_{y} = \frac{1}{3}\left(9 + 4 + 25\right) \approx 12.6
$$

```{r}
y <- c(2, 3, 10)
mean(y)
var(y) # Up to the scaling factor
```
