---
title: "ARIMA: GDP example"
author: "Boyko Amarov"
date: "12/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(xts)
library(tidyverse)

dt <- read_csv("https://raw.githubusercontent.com/feb-uni-sofia/timeseries-2021-2022/main/data/gdp_bg_qrt_2000-2017.csv")
gdp <- xts(dt$GDP, order.by = as.yearqtr(dt$Index, format = "%YQ%q"))
```

```{r}
plot(gdp)
```

## Growth series 1 (quarter to previous quarter)

For a series $y_1, \ldots, y_T$ the growth rate is given by

$$
g_t = \frac{y_t - y_{t - 1}}{y_{t - 1}} \approx \log(y_t) - \log(y_{t - 1})
$$

```{r}
gs_1 <- log(gdp) - log(stats::lag(gdp, k = 1))
plot(gs_1)
```

This series still shows strong seasonality. We would like to avoid that for now, as our
familiar arima models cannot handle seasonal effects (we will discuss this in the following weeks).
Let us define the growth rate in a different way by comparing the GDP of a quarter with the GDP of the same quarter during the previous year.

$$
g^{(4)}_{t} = \log(y_t) - \log(y_{t - 4})
$$
```{r}
gs4 <- log(gdp) - log(stats::lag(gdp, k = 4))
plot(gs4)
```
```{r}
acf(gs4, na.action = na.pass)
pacf(gs4, na.action = na.pass)
```
```{r}
fit_AR1 <- arima(gs4, order = c(1, 0, 0))
fit_AR1
```

## Using the `forecast` package

```{r}
library(forecast)
fit_AR1 <- Arima(gs4, order = c(1, 0, 0))
```

```{r}
autoplot(forecast(fit_AR1))
```

$$
\hat{y}_{T + 1|T}\\
\hat{y}_{T + 2|T}
$$


```{r}
predict(fit_AR1, n.ahead = 40)
```

## Prediction with ARMA processes

Let $y_t$ follow a stationary ARMA process. The Wold decomposition states that every stationary ARMA process can be represented as MA($\infty$).

$$
\begin{align}
y_t = \mu + \sum_{j = 0}^{\infty} \psi_j e_{t - j}, \quad e_t \sim WN(\sigma^2)
\end{align}
$$

with the restriction that $\psi_0 = 1$ and that the square sum of the coefficients is finite:

$$
\sum_{j = 0}^{\infty} \psi_j^2 \le \infty
$$

The latter condition ensures that the process has a finite variance. Let

$$
\hat{y}_t = \mu + \sum_{k = 0}^{\infty} \theta_k u_{t - k}
$$

be a forecast function. The error between the value of the process and the prediction for a $h$ periods ahead prediction is:
$$
y_{t + h}- \hat{y}_{t + h|t}
$$

it can be shown, that the forecast function with the smallest expected squared error is the conditional expectation of $y_{t + h}$ given the information up to time $t$.

