---
title: "ARIMA: GDP example"
author: "Boyko Amarov"
date: "12/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(xts)
library(tidyverse)
library(fpp)
# install.packages(c("fpp", "forecast"))

data(elecequip)
dt <- read_csv("https://raw.githubusercontent.com/feb-uni-sofia/timeseries-2021-2022/main/data/gdp_bg_qrt_2000-2017.csv")

gdp <- xts(dt$GDP, order.by = as.yearqtr(dt$Index, format = "%YQ%q"))
```

```{r}
plot(gdp)
```

## Growth series 1 (quarter to previous quarter)

For a series $y_1, \ldots, y_T$ the growth rate is given by

$$
g_t = \frac{y_t - y_{t - 1}}{y_{t - 1}} \approx \log(y_t) - \log(y_{t - 1})
$$

```{r}
gs_1 <- log(gdp) - log(stats::lag(gdp, k = 1))
plot(gs_1)
```

This series still shows strong seasonality. We would like to avoid that for now, as our
familiar arima models cannot handle seasonal effects (we will discuss this in the following weeks).
Let us define the growth rate in a different way by comparing the GDP of a quarter with the GDP of the same quarter during the previous year.

$$
g^{(4)}_{t} = \log(y_t) - \log(y_{t - 4})
$$
```{r}
gs4 <- log(gdp) - log(stats::lag(gdp, k = 4))
plot(gs4)
```
```{r}
acf(gs4, na.action = na.pass)
pacf(gs4, na.action = na.pass)
```
```{r}
fit_AR1 <- arima(gs4, order = c(1, 0, 0))
fit_AR1
```

## Using the `forecast` package

```{r}
library(forecast)
fit_AR1 <- Arima(gs4, order = c(1, 0, 0))
```

```{r}
autoplot(forecast(fit_AR1))
```

$$
\hat{y}_{T + 1|T}\\
\hat{y}_{T + 2|T}
$$


```{r}
predict(fit_AR1, n.ahead = 40)
```
$$
y_{t} = \alpha y_{t -1} + e_t\\
Ey_{t} = \alpha E y_{t -1}\\
Ey_{t} =  \hat{y}_t = \hat{\alpha} \hat{y}_{t - 1}\\
$$


