---
title: "DF-test"
author: "Boyko Amarov"
date: "1/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(urca)
## install.packages("urca")
```


Random walk model without a drift

$$
y_{t} = y_{t - 1} + u_{t}, u_t \sim WN(\sigma^2)
$$

$$
\lambda - 1 = 0 \implies 
\lambda^* = 1
$$

The solution of this characteristic equation is 1. We say that the characteristic equation has a unit root.

## Random walks simulation 

```{r}
rw1 <- arima.sim(model = list(order = c(0, 1, 0)), n = 200)
rw2 <- arima.sim(model = list(order = c(0, 1, 0)), n = 200)

# rw2 <- 1.5 * rw1 + rnorm(n = 200)
plot(rw1)
plot(rw2)
```

```{r}
summary(lm(rw2 ~ rw1))
```
Regression model
$$
y_{t}: \text{ random walk 2}\\
x_{t}: \text{ random walk 1}\\
y_{t} = \beta_0 + \beta_1 x_{t} + e_{t}, e_{t} \sim N(0, \sigma^2)
$$
Estimated regression equation

$$
\hat{y}_{t} = 6.15 - 1.78 x_{t}
$$
T-Test of the hypothesis 

$$
H_0: \beta_1 = 0\\
H_1: \beta_1 \neq 0
$$
We reject the null hypothesis if the p-value of the test is less than 0.05 (convention).

spurious regression



$$
\hat{y}_{t} = -1.86 + 0.19 x_{t}
$$


# Unit root tests

AR(1) process without a constant

$$
y_{t} = \phi_1 y_{t - 1} + u_{t}, u_t \sim WN(\sigma^2)
$$
If the process has a unit root, it follows that $\phi_1 = 1$.

We would like to test the hypothesis:

$$
H_0: \phi_1 = 1\\
H_1: \phi < 1
$$
First difference
$$
\Delta y_{t} = y_{t} - y_{t - 1}
$$

First difference of a random walk

$$
y_{t} = y_{t - 1} + u_{t}\\
y_{t} - y_{t - 1} = u_{t}\\
\Delta y_{t} = u_{t}
$$

The first difference of the random walk process is stationary, because it is 
simply equal to the pure random process $u_{t}$ and it is stationary by the properties of the pure random process. We say that the process (in this case the random walk process) is integrated of order 1.

## Dickey-Fuller Unit Root Test

$$
y_{t} - y_{t - 1} = \phi_1 y_{t - 1} + u_{t} - y_{t - 1} \\
\iff\\
\Delta y_{t} = (\phi_1 - 1)y_{t - 1} + u_t
$$
$$
\phi_1 - 1 = 0 \iff \phi_1 = 1
$$

We can use the t-statistic for testing the hypothesis that the coefficient 
of $y_{t - 1}$ equals 0, but the distribution of this statistic under the $H_0$ (assuming that $\phi_1 = 1$) is not a t-distribution. Therefore we need to use 
special critical values in order to perform the test.




## Performing the DF-Test: AR(1)

```{r}
## Unit root (ur)
summary(ur.df(rw1, type = "none", lags = 0))
```
-1 in the regression formula instructs lm to exclude the
intercept from the regression equation.

$$
\underbrace{\Delta y_{t}}_{z.diff} = \underbrace{y_{t - 1}}_{z.lag.1} + e_t
$$
t-value is the value of the t-statistic.

$$
t = \frac{\hat{\beta}_1 - 0}{SE(\hat{\beta}_1)}
$$
$$
-0.007011 / 0.008509 = -0.824
$$
To perform the test, compare the t-value with the critical values:

      1pct  5pct 10pct
tau1 -2.58 -1.95 -1.62

and we reject the null hypothesis if the t-value value is lower than the critical values.

-0.824 > -1.62 (the critical value corresponding to a 10% significance level), so we cannot reject the null hypothesis.

$$
H_0: \phi_1 - 1 = 0
$$


## Simulation 2 (Stationary AR(1))

$$
y_{t} = 100 + 0.7y_{t - 1} + u_{t}, u_t \sim WN(1)
$$

```{r}
set.seed(5454) ## For reproducibility of the simulation results
ar1 <- 100 + arima.sim(model = list(order = c(1, 0, 0), ar = c(0.7)), n = 200)
plot(ar1)
```

```{r}
summary(ur.df(ar1, type = "none", lags = 0))
```
```{r}
summary(ur.df(ar1, type = "drift", lags = 0))
```
Compare the t-value (-6.253) to the critical values:

      1pct  5pct 10pct
tau2 -3.46 -2.88 -2.57

-6.253 < -3.46 (1-percent critical value) => we reject the
null hypothesis at a 1-percent significance level.


## Simulation 3 (Stationary AR(1) with trend)

$$
y_{t} = 10 + 0.1t + 0.7 y_{t - 1} + u_{t}
$$

```{r}
set.seed(5454) ## For reproducibility of the simulation results
ar2 <- 10 + 0.1*(1:200) + arima.sim(model = list(order = c(1, 0, 0), ar = c(0.7)), n = 200)
plot(ar2)
```

```{r}
summary(ur.df(ar2, type = "drift", lags = 0))
```
-0.9984 > -2.57 => the test fails to reject the null hypothesis
at any reasonable significance level.

```{r}
summary(ur.df(ar2, type = "trend", lags = 0))
```
      1pct  5pct 10pct
tau3 -3.99 -3.43 -3.13

-6.2976 < -3.99 => reject the null hypothesis at a 1-percent significance level.







