---
title: "ARIMA: GDP example"
author: "Boyko Amarov"
date: "12/21/2021"
output: html_document
---

```{r setup}
library(tidyverse)
library(xts)
library(forecast)
library(fpp)

data(elecequip)
elecadj <- elecequip %>%  stl(s.window='periodic') %>% seasadj()
```
```{r}
plot(elecadj)
```

## Information Criteria

As the model complexity increases, e.g. AR(2) is more complex than AR(1) because it has more parameters and is therefore more flexible, the model can fit more closely to the data.
However, this reduces the effective information used to estimate each parameter of the model (limited information spread over a larger number of coefficients). Furthermore, more complex models
tend to fit the noise in the data rather than to capture a systematic development over time (i.e. signal). This is where information criteria come to help choose a model that balances complexity
versus close fit to the observations.

$$
\text{AIC} = -2\log(\text{Likelihood}) + 2(p + q + k + 1)\\
k = \begin{cases}
1 & \delta \neq 0\\
0 & \delta = 0
\end{cases}
$$

In the above equation $k$ equals one or zero depending on whether the model contains a constant ($\delta$), $p$ and $q$ are the orders of the AR and MV parts, and one is added to the penalty because of the estimation of the $\sigma$ parameter (the standard deviation of the white noise process).

Another commonly used information criterion is the Bayes information criterion (BIC):

$$
\text{BIC} = AIC + (\log(T) - 2)(p + q + k + 1)
$$

