---
title: "DF-test"
author: "Boyko Amarov"
date: "1/4/2022"
output: html_document
---

ARIMA(p, d, q)


```{r}
library(urca)
```


Random walk model without a drift

$$
y_{t} = y_{t - 1} + e_{t}, e_t \sim WN(\sigma^2)\\
\Delta y_{t} = y_{t} - y_{t - 1} = e_{t}\\
f' = \lim_{\epsilon \to 0}\frac{f(x) - f(x + \epsilon)}{\epsilon}
$$

$$
\lambda - 1 = 0 \implies 
\lambda^* = 1
$$

The solution of this characteristic equation is 1. We say that the characteristic equation has a unit root.

ARIMA(0, 1, 0)

## Random walks simulation 

```{r}
rw1 <- arima.sim(model = list(order = c(0, 1, 0)), n = 200)
rw2 <- arima.sim(model = list(order = c(0, 1, 0)), n = 200)

# rw2 <- 1.5 * rw1 + rnorm(n = 200)
plot(rw1)
plot(rw2)
```


```{r}
summary(lm(rw2 ~ rw1))
```

Regression model

$$
y_{t}: \text{ random walk 2}\\
x_{t}: \text{ random walk 1}\\
y_{t} = \beta_0 + \beta_1 x_{t} + e_{t}, e_{t} \sim N(0, \sigma^2)
$$
Estimated regression equation

$$
\hat{y}_{t} = 6.15 - 1.78 x_{t}
$$
T-Test of the hypothesis 

$$
H_0: \beta_1 = 0\\
H_1: \beta_1 \neq 0
$$
We reject the null hypothesis if the p-value of the test is less than 0.05 (convention).

spurious regression

$$
\hat{y}_{t} = -1.86 + 0.19 x_{t}
$$


# Unit root tests

AR(1) process without a constant

$$
y_{t} = \alpha_1 y_{t - 1} + e_{t}, e_t \sim WN(\sigma^2)
$$
If the process has a unit root, it follows that $\alpha_1 = 1$.

We would like to test the hypothesis:

$$
H_0: \alpha_1 = 1\\
H_1: \alpha_1 < 1
$$

The first difference of the random walk process is stationary, because it is
simply equal to the pure random process $e_{t}$ and it is stationary by the properties of the pure random process. We say that the process (in this case the random walk process) is integrated of order 1.

## Dickey-Fuller Unit Root Test

$$
y_{t} - y_{t - 1} = \alpha_1 y_{t - 1} + e_{t} - y_{t - 1} \\
\iff\\
\Delta y_{t} = (\alpha_1 - 1)y_{t - 1} + e_t
$$
$$
\alpha_1 - 1 = 0 \iff \alpha_1 = 1
$$

We can use the t-statistic for testing the hypothesis that the coefficient 
of $y_{t - 1}$ equals 0, but the distribution of this statistic under the $H_0$ (assuming that $\alpha_1 = 1$) is not a t-distribution. Therefore, we need to use
special critical values in order to perform the test.

```{r}
summary(ur.df(rw1, type = "none", lags = 0))
```
-1 in the regression formula instructs lm to exclude the
intercept from the regression equation.

$$
\underbrace{\Delta y_{t}}_{z.diff} = \underbrace{y_{t - 1}}_{z.lag.1} + e_t
$$
t-value is the value of the t-statistic.

$$
t = \frac{\hat{\beta}_1 - 0}{SE(\hat{\beta}_1)}
$$
$$
-0.007011 / 0.008509 = -0.824
$$
To perform the test, compare the t-value with the critical values:

      1pct  5pct 10pct
tau1 -2.58 -1.95 -1.62

and we reject the null hypothesis if the t-value value is lower than the critical values.

-0.824 > -1.62 (the critical value corresponding to a 10% significance level), so we cannot reject the null hypothesis.

$$
H_0: \alpha_1 - 1 = 0
$$

## Simulation 2: Stationary AR(1) with zero level

$$
y_{t} = 0.7y_{t - 1} + e_{t}, e_t \sim WN(1)
$$

```{r}
set.seed(5445)
ar1 <- arima.sim(n = 200, model = list(ar = 0.7))
plot(ar1)
```

```{r}
summary(ur.df(ar1, type = "none", lags = 0))
```

-6.54 < -2.58 < -1.95 => reject the null hypothesis of presence of a unit root

## Simulation 2: Stationary AR(1) with non-zero level

$$
y_{t} = 100 + 0.7y_{t - 1} + e_{t}, e_t \sim WN(1)
$$

```{r}
ar2 <- 100 + ar1
plot(ar2)
```

```{r}
summary(ur.df(ar2, type = "none", lags = 0))
```
```{r}
summary(ur.df(ar2, type = "drift", lags = 0))
```
Compare the t-value (-6.253) to the critical values:

      1pct  5pct 10pct
tau2 -3.46 -2.88 -2.57

-6.253 < -3.46 (1-percent critical value) => we reject the
null hypothesis at a 1-percent significance level.


## Simulation 3 (Stationary AR(1) with trend)

$$
y_{t} = 10 + 0.1t + 0.7 y_{t - 1} + e_{t}
$$


```{r}
ar3 <- 10 + 0.1*(1:length(ar1)) + ar1
plot(ar3)
```

```{r}
summary(ur.df(ar3, type = "drift", lags = 0))
```
-0.9984 > -2.57 => the test fails to reject the null hypothesis
at any reasonable significance level.

```{r}
summary(ur.df(ar3, type = "trend", lags = 0))
```
      1pct  5pct 10pct
tau3 -3.99 -3.43 -3.13

-6.2976 < -3.99 => reject the null hypothesis at a 1-percent significance level.

## ADF-Test

Let us consider an AR(2) process

$$
y_{t} = \alpha_1 y_{t - 1} + \alpha_2 y_{t - 2} + e_t
$$

If this process has a unit root, then its characteristic equation

$$
\lambda ^2 - \alpha_1 \lambda - \alpha_2 = 0
$$

has a solution $\lambda^{*} = 1$, therefore, setting $\lambda = 1$ satisfies the equation, and we obtain:

$$
1^2 - \alpha_1 - \alpha_2 = 0
$$

Therefore, if the process has a unit root, its coefficients sum to one. This is a hypothesis that we can test using data. Following the approach of the DF-Test, we can try to construct a regression equation where one of the coefficients is $1 - \alpha_1 - \alpha_2$. Then we can conduct a test of the null hypothesis that this coefficient is zero.

It turns out that we can reparemetrize the AR(2) model to obtain such an equation:

$$
\begin{align}
y_{t} - y_{t - 1} & = \alpha_1 y_{t - 1} - y_{t - 1} + \alpha_2 y_{t - 2} + e_t\\
\Delta y_{t} & = (\alpha_1 - 1)y_{t - 1} + \alpha_2 y_{t - 2} + e_t\\
\Delta y_{t} & = (\alpha_1 - 1)y_{t - 1} + \alpha_2 y_{t - 1} - \alpha_2 y_{t - 1} + \alpha_2 y_{t - 2} + e_t\\
\Delta y_{t} & = (\alpha_1 + \alpha_2 - 1)y_{t - 1} - \alpha_2 y_{t - 1} \alpha_2 y_{t - 2} + e_t\\
\Delta y_{t} & = (\alpha_1 + \alpha_2 - 1)y_{t - 1} - \alpha_2(y_{t - 1} - y_{t - 2}) + e_t\\
\Delta y_{t} & = (\alpha_1 + \alpha_2 - 1)y_{t - 1} - \alpha_2\Delta y_{t - 1} + e_t\\
\end{align}
$$

In the general case of an AR(p) process the equation looks more complicated but it is derived in exactly the same way as for the AR(2) process:

$$
\Delta y_t = (\rho - 1) y_{t - 1} + \theta_{1} \Delta y_{t - 1} + \theta_2 \Delta y_{t - 2} + \ldots + \theta_{p - 1} \Delta y_{t - p + 1} + e_t
$$

where $\rho$ and $\theta_i$ are defined as:

$$
\rho = \sum_{j = 1}^p \alpha_j \\
\theta_i = - \sum_{j = i + 1}^{p} \alpha_j
$$

## Simulation: AR(3)

$$
y_t = 0.6y_{t - 1} + 0.4y_{t - 2} - 0.7y_{t - 2} + e_t, e_t \sim WN(1)
$$

```{r}
set.seed(521)
ar4 <- arima.sim(n = 200, model = list(ar = c(0.6, 0.4, -0.7)))
plot(ar4)
```
```{r}
summary(ur.df(ar4, type = "none", lags = 0))
```
```{r}
summary(ur.df(ar4, type = "none", selectlags = "AIC"))
```

## KPSS Test

The KPSS test considers stationarity as the null hypothesis. Let's consider the following process:

$$
y_t = \alpha_t + \beta t + e_t, e_t \sim WN(\sigma^2)
$$

where $\alpha_t$ is a random walk process: $\alpha_t = \alpha_{t - 1} + \epsilon_t$. If $y_t$ is trend-stationary (i.e. has no unit roots), then the variance of $\epsilon_t$ is zero (i.e. $\alpha_t$ is a constant process).

With the KPSS test we reject the null hypothesis (trend-stationarity)
for _large_ values of the test statistic.

```{r}
summary(ur.kpss(rw1))
summary(ur.kpss(ar1))
```
